---
title: "2.Seminar Part II - Peak-gene Linkages"
author: 
- name: "Maria Lucia Romero Rivero"
  email: MLROMEROR@clinic.cat
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
    fig_captions: yes

---


# Introduction

Hi to everyone!
In this notebook we are going to practice how to perform the Peak-gene linkage analysis using B-cell of tonsil multiome dataset (ATAC + GEX).This analysis was developed in the `Signac`( _Stuart et al._ , 2020) package by 10x Genomics.


In this report we will continue working with the previous Seurat object saved as `3.tonsil_bcelling.rds`in the results/R_object directory.  

To do this we must first download the following files:



<details>
  <summary>**Load packages**</summary>
  

```{r eval=FALSE}
BiocManager::install("Repitools")

```


```{r setup, message=FALSE, warning=FALSE}
library(Signac)
library(Seurat)
library(ggplot2)
library(ggpubr)
library(tidyverse)
library(EnsDb.Hsapiens.v86)
library(plyr)
library(reshape2)
library(data.table)
library(GenomicRanges)
library(harmony)
library(hdf5r)
library(stringr)
library(ggpubr)
library(RColorBrewer)
library(magick)
library(knitr) 
library(biovizBase)
library(patchwork)
library(Repitools)
library(qlcMatrix)
library(BSgenome.Hsapiens.UCSC.hg38)
set.seed(123)
```
</details>

  
# Loading data
 
## Directory path 

```{r}
# Paths
path_to_data <- here::here("results/R_objects/")
path_to_save <- here::here("results/R_objects/")
```


```{r message=FALSE, warning=FALSE}
tonsil_bcell <- readRDS(paste0(path_to_data,"3.tonsil_bcell.rds"))
```

# UMAP 

```{r}
 DimPlot(
    tonsil_bcell,
    group.by = "seurat_clusters_name",
    reduction = "wnn.umap",
    pt.size = 0.1, label = T
  )
```


# Link peaks

```{r eval=FALSE, include=FALSE}
DefaultAssay(tonsil_bcell) <- "ATAC"

# first compute the GC content for each peak
tonsil_bcell <- RegionStats(tonsil_bcell, genome = BSgenome.Hsapiens.UCSC.hg38)

# link peaks to all genes
tonsil_bcell <- LinkPeaks(
  object = tonsil_bcell,
  peak.assay = "ATAC",
  expression.assay = "RNA"
)
```

```{r eval=FALSE, include=FALSE}
DefaultAssay(tonsil_bcell) <- "ATAC"

# first compute the GC content for each peak
tonsil_bcell <- RegionStats(tonsil_bcell, genome = BSgenome.Hsapiens.UCSC.hg38)

# link peaks to  BCL6 and PRDM1 genes at 2e+8 TSS 
tonsil_bcell <- LinkPeaks(
  object = tonsil_bcell,
  peak.assay = "ATAC",
  expression.assay = "RNA",
  genes.use = c("BCL6","PRDM1"),
  distance = 2e+8
)
```

```{r}
DefaultAssay(tonsil_bcell) <- "ATAC"

# first compute the GC content for each peak
tonsil_bcell <- RegionStats(tonsil_bcell, genome = BSgenome.Hsapiens.UCSC.hg38)

# link peaks to genes
tonsil_bcell <- LinkPeaks(
  object = tonsil_bcell,
  peak.assay = "ATAC",
  expression.assay = "RNA",
  genes.use = c("BCL6","PRDM1"),
  distance = 1e+9
)


```


We are going to create a dataframe with all the links related to PRDM1 and BCL6 genes.

```{r}
df_LinkPeaks<-as.data.frame(Links(tonsil_bcell))
head(df_LinkPeaks[df_LinkPeaks$gene=="BCL6",])
```


```{r}
head(df_LinkPeaks[df_LinkPeaks$gene=="PRDM1",])
```


```{r}
head(df_LinkPeaks)
```



```{r}
idents.plot<-c("GCBC","NBC","MBC", "PC")

DefaultAssay(tonsil_bcell) <- "ATAC"

CoveragePlot(
object = tonsil_bcell,
region = "BCL6",
features = "BCL6",
expression.assay = "RNA",
idents = idents.plot,
extend.upstream = 5000000,
extend.downstream = 10000000
#tile = TRUE
)

```


```{r}
CoveragePlot(
object = tonsil_bcell,
region = "PRDM1",
features = "PRDM1",
expression.assay = "RNA",
idents = idents.plot,
extend.upstream = 10000000,
extend.downstream = 10000000
#tile = TRUE
)

```


# Modality Weights

```{r}
rna.weight<- VlnPlot(tonsil_bcell, features = "RNA.weight", group.by = "seurat_clusters_name",pt.size = 0)
atac.weight<- VlnPlot(tonsil_bcell, features ="ATAC.weight", group.by = "seurat_clusters_name",pt.size = 0)

rna.weight + ggtitle("RNA modality weight")
atac.weight+ ggtitle("ATAC modality weight")
```



# Save


```{r}
saveRDS(tonsil_bcell,paste0(path_to_save,"4.tonsil_bcell_linkpeaks.rds"))
```

<details>
  <summary>**Session Info**</summary>
  
```{r}
sessionInfo()
```


</details>